\large{\bf Normality and large effect loci in quantitative genetics }

### 2023-07-04 writeup.tex
.
low mem plotting

%%%% LAST VERSION
% 1. *Intro*: Motivation about how many results about quant gen and methods assume Gaussian distributions.
%     Mention Levy walk macroevolution papers.
%     Point out that usual quant gen has *no* sweeps, while stable distributions would have sweeps (since some s would be bigger than 1/N).
%     Motivate with mixture model of effect sizes.
%     If stable distributions are a better model, what does that tell us? We can better understand the *noise* in evolution, i.e., predictability and saltation-ness.
%     (PETER)
% 
% 2. Motivation: empirical SNPs distributions in real data (PETER)
% 
% 3. Introducing infinitesimal model (in this case) (TODD)
% 
% 4. Proof following BEV in stable case (TODD)
% 
% 5. Theoretical results on stationary distributions (TODD)
% 
% 6. Simulations (PETER)

\title{\large{\bf
    Abnormal quantitative genetics?
}}

@@

\comment{remove the following?}
It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Convergence to the non-infinitesimal model}  % TODD

%%%%%%%%%%%%%%%%%%%%%%
\section{Stationary distributions}  % TODD

%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations}  % PETER

@@

We compared two situations:
(1) the effect size distribution was Gaussian with mean zero and standard deviation $\sigma_g$,
and $f(z) = e^{\alpha z^2}$; and
(2) the effect size distribution was Cauchy, centered at zero and with width $\delta - \gamma$
and $f(z) = (\delta/\gamma) (\gamma + z^2) / (\delta + z^2)$.
\comment{TODO: just refer to equations above?}
We ran simulations of each with $N=XXX$ for XXX time steps of ``burn-in''
(judged to be long enough as the number of segregating mutations had plateaued),
following which we recorded the phenotypes of each new offspring and their two parents
for XXX time steps (a total of around YYY offspring).

Figure XXX shows the resulting distributions of segregation noise:
i.e., the distribution of child phenotype minus average of parents' phenotypes,
across all offspring following burn-in.

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_trait_traces}
    \end{center}
    \caption{
        Median trait values over time \textbf{(A and C)}
        and distributions of final trait values \textbf{(B and D)}
        in neutral simulations
        with Normal \textbf{(A and B)} and Cauchy \textbf{(C and D)}
        effect size distributions.
        Simulations were begun with no genetic variation,
        had populations of size $N=1000$,
        a genome of length $L=10^8$ and mutation rate of $\mu=10^{-9}$;
        the effect size distributions were
        Normal with mean 0 and standard deviation $1/\sqrt{4NL\mu}$
        and Cauchy with center 0 and scale $1/(4NL\mu)$, respectively.
        \label{fig:trait_distrns}
    }
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_trait_traces}
    \end{center}
    \caption{
        Median trait values over time \textbf{(A and C)}
        and distributions of final trait values \textbf{(B and D)}
        in neutral simulations
        with Normal \textbf{(A and B)} and Cauchy \textbf{(C and D)}
        effect size distributions.
        Simulations were begun with no genetic variation,
        had populations of size $N=1000$,
        a genome of length $L=10^8$ and mutation rate of $\mu=10^{-9}$;
        the effect size distributions were
        Normal with mean 0 and standard deviation $1/\sqrt{4NL\mu}$
        and Cauchy with center 0 and scale $1/(4NL\mu)$, respectively.
        \label{fig:trait_distrns}
    }
\end{figure}

>>>>

\title{\large{\bf
    Normality and large effect loci in quantitative genetics
}}

@@

\comment{remove or shorten the following?}
It is clear in the Gaussian case
that $U^{[1]}$ and $V^{[1]}$ are asymptotically independent.
More generally, independence depends on how the model is set up.
If the allelic effects $\eta_\ell^{[i]}$ are i.i.d.,
then even for finite $M$, the parental contributions
$(U_M^{[1]}, V_M^{[1]}, U_M^{[2]}, V_M^{[2]})$ are independent.
If the distribution of $\eta_\ell^{[i]}$ is chosen to obtain a non-Gaussian limit,
then Lemma~\ref{lem:parental_contribs} implies that $R$ cannot be independent of the parental traits.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_trait_traces}
    \end{center}
    \caption{
        Median trait values over time \textbf{(A and C)}
        and distributions of final trait values \textbf{(B and D)}
        in neutral simulations
        with Gaussian \textbf{(A and B)} and Cauchy \textbf{(C and D)}
        effect size distributions.
        Simulations were begun with no genetic variation,
        had populations of size $N=1000$,
        a genome of length $L=10^8$ and mutation rate of $\mu=10^{-9}$;
        the effect size distributions were
        Normal with mean 0 and standard deviation $1/\sqrt{4NL\mu}$
        and Cauchy with center 0 and scale $1/(4NL\mu)$, respectively.
        \label{fig:trait_distrns}
    }
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    Segregation noise with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
    from the same simulations shown in Figure~\ref{fig:trait_distrns},
    for all individuals born during the last 1,000 time units of the simulation.
    \textbf{(A,D)} Segregation noise (offspring trait minus midparent) plotted against midparent.
    \textbf{(B,E)} Histograms of segregation noise.
    \textbf{(B,E)} Enrichment in quantiles of segregation noise conditional on midparent value.
    To compute the latter,
    let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of segregation noise
    into 50 bins with roughly equal numbers in each (i.e., the quantiles),
    let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
    is in $[a, b)$ whose segregation noise is in $[q_{k-1}, q_k)$,
    and let $p_k$ be the proportion of all offspring with segreation noise in $[q_{k-1}, q_k)$
    (so, $p_k \approx 0.02$).
    Then, each line shows $p_k(a,b)/p_k$
    plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
    for $(a,b)$ chosen to span the 5\% of midparents centered on:
    (red) the 10\% quantile,
    (black) the median, and
    (green) the 90\% quantile of midparent value.
    \caption{
        \label{fig:seg_noise}
    }
\end{figure}

We first consider the neutral case (i.e., with $\mu \equiv 0$).
As shown in Figure~\ref{fig:trait_distrns},
a Cauchy effect size distribution, unsurprisingly,
produces a distribution of trait values with more extreme values
and a rougher path of population median trait value over time
than in the Gaussian case.
Indeed, the sum of all effects along the path from the start of the simulation
to an arbitrary individual will approximate either Brownian motion (in the Gaussian case)
or a Cauchy process (in the Cauchy case).
Figure~\ref{fig:seg_noise} shows how segregation noise
depends on midparent value.
In the Gaussian case, segregation noise is Gaussian and independent of midparent value,
as shown by the flat lines at 1 in Figure~\ref{fig:seg_noise}C. 
On the other hand, in the Cauchy case we can see from either Figure~\ref{fig:seg_noise}D or F
that segregation noise depends strongly on midparent trait,
as we would expect in the presence of several large-effect alleles.

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_trait_traces}
    \end{center}
    \caption{
        As in Figure~\ref{fig:trait_distrns},
        except under a model of stabilizing selection (see text for details).
        \label{fig:sel_trait_distrns}
    }
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_seg_noise}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except under a model of stabilizing selection (see text for details).
        \label{fig:sel_seg_noise}
    }
\end{figure}

We next compared two situations of stabilizing selection, motivated by the results above
\comment{TODO: refer to where?}:
(1) the effect size distribution was Gaussian with mean zero and standard deviation $\sigma_g$,
and $f(z) = e^{\alpha z^2}$; and
(2) the effect size distribution was Cauchy, centered at zero and with width $\delta - \gamma$
and $f(z) = (\delta/\gamma) (\gamma + z^2) / (\delta + z^2)$.
Figure XXX shows the resulting distributions of segregation noise:
i.e., the distribution of child phenotype minus average of parents' phenotypes,
across all offspring following burn-in.
